\vspace{-2.5em} Excitation and Detection Efficiency Under Scanned Light Sheet Illumination \vspace{-2.5em}

### 2017-06-26 notes/2017-06-21-light-sheet-illumination/report/report.tex
Revised light sheet illumination notes with quantified error.

\section{Scanned Beam Excitation Efficiency}
We define the excitation efficiency of a fluorophore as the fraction of 
incident power that excites the fluorophore. If a fluorophore with absorption
dipole moment $\bs{\mu}_{\text{abs}}$ is placed in a time-independent complex
electric field $\mb{E}$, then the excitation efficiency is given by
\begin{align}
  \eta_{\text{exc}}(x,y,z) = \frac{|\bs{\mu}_{\text{abs}} \cdot \mb{E}(x,y,z)|^2}{|\mb{E}(x,y,z) |^2}.
\end{align}
If the fluorophore is placed in the path of a scanned laser then the electric
field becomes time dependent. If the laser is scanned extremely fast we would
need to consider the coherence of the electric field, but we will only consider
slow scanning here to simplify the calculation. Specifically, we require that
$v \ll w_0/\tau_c$---the scan velocity must be much less than the beam width
divided by the coherence time. In this case the excitation efficiency is
\begin{align}
  \eta_{\text{exc}}(x,y,z) = \frac{\intinf|\bs{\mu}_{\text{abs}} \cdot \mb{E}(x,y,z,t)|^2dt}{\intinf|\mb{E}(x,y,z,t)|^2dt}\label{eq:excitation}
\end{align}
We plug equation \ref{eq:scanned} into equation \ref{eq:excitation}, express
$\bs{\mu}_{\text{abs}}$ in spherical coordinates
\begin{align}
  \bs{\mu}_{\text{abs}} = \cos\Phi\sin\Theta\mh{x} + \sin\Phi\sin\Theta\mh{y} + \cos\Theta\mh{z},
\end{align}
and evaluate the integrals (see Appendix for details) to write the
excitation efficiency as
\begin{align}
  \eta_{\text{exc}}(x,y,z) = \frac{\sin^2\Theta\cos^2(\Phi-\phi_{\text{pol}}) + \cos^2\Theta\frac{x^2\cos^2\phi_{\text{pol}} + \frac{1}{4}w^2(z)\sin^2\phi_{\text{pol}}}{z_0^2}}{1+\frac{x^2\cos^2\phi_{\text{pol}} + \frac{1}{4}w^2(z)\sin^2\phi_{\text{pol}}}{z_0^2}}.\label{eq:strong}
\end{align}
If the beam is weakly focused then we can ignore the longitudinal excitation and
the excitation efficiency simplifies to
\begin{align}
  \eta_{\text{exc}} = \sin^2\Theta\cos^2(\Phi-\phi_{\text{pol}}).
\end{align}

@@

To ignore the longitudinal component, we require that the ratio of the beam
waist to the Rayleigh range be much less than 2 ($w/z_0 \ll 2$).  The excitation
beam in \cite{wu2013} satisfies this criteria
$w/z_0 = 1.2\ \mu\text{m}/9\ \mu\text{m} = 0.13$, but we'll need to be careful if
we want to use beams that are more strongly focused.

>>>>

\section{Scanned Beam Excitation Efficiency}
We define the excitation efficiency of a fluorophore as the fraction of 
incident power that excites the fluorophore. If a fluorophore with absorption
dipole moment $\bs{\mu}_{\text{abs}}$ is placed in a time-independent complex
electric field $\mb{E}$, then the excitation efficiency is given by
\begin{align}
  \eta_{\text{exc}} = \frac{|\bs{\mu}_{\text{abs}} \cdot \mb{E}(x,y,z)|^2}{|\mb{E}(x,y,z) |^2}.
\end{align}
If the fluorophore is placed in the path of a scanned laser then the electric
field becomes time dependent. If the laser is scanned quickly we would need to
consider the coherence of the electric field, but we will only consider slow
scanning here to simplify the calculation. Specifically, we require that
$v \ll w_0/\tau_c$---the scan velocity must be much less than the beam width
divided by the coherence time. In this case the excitation efficiency is
\begin{align}
  \eta_{\text{exc}} = \frac{\intinf|\bs{\mu}_{\text{abs}} \cdot \mb{E}(x,y,z,t)|^2dt}{\intinf|\mb{E}(x,y,z,t)|^2dt}\label{eq:excitationx}
\end{align}
We plug equation \ref{eq:scanned_e} into equation \ref{eq:excitationx}, express
$\bs{\mu}_{\text{abs}}$ in spherical coordinates
\begin{align}
  \bs{\mu}_{\text{abs}} = \cos\Phi\sin\Theta\mh{x} + \sin\Phi\sin\Theta\mh{y} + \cos\Theta\mh{z},
\end{align}
and evaluate the integrals (see Appendix for details) to write the
excitation efficiency as
\begin{align}
  \eta_{\text{exc}} = \frac{\sin^2\Theta\cos^2(\Phi-\phi_{\text{pol}}) + \cos^2\Theta\frac{x^2\cos^2\phi_{\text{pol}} + \frac{1}{4}w^2(z)\sin^2\phi_{\text{pol}}}{z_0^2}}{1+\frac{x^2\cos^2\phi_{\text{pol}} + \frac{1}{4}w^2(z)\sin^2\phi_{\text{pol}}}{z_0^2}}.\label{eq:strong}
\end{align}
If the beam is weakly focused then we can ignore the longitudinal excitation and
the excitation efficiency simplifies to
\begin{align}
  \eta_{\text{exc}} \approx \sin^2\Theta\cos^2(\Phi-\phi_{\text{pol}}).\label{eq:weak_approx}
\end{align}

\textbf{How good is the approximation in equation \ref{eq:weak_approx}?}

For longitudinal fluorophores equation \ref{eq:weak_approx} is a very bad
approximation---it predicts that a longitudinal fluorophore will not be excited
at all. We can't use a percentage error because the approximation completely
ignores the excitation of longitudinal dipoles.

Instead, we can compare the size of signals from longitudinal and transverse
fluorophores. If the signal from longitudinal fluorophores is less than the
signal from noise and background, then we can ignore the signal from
longitudinal fluorophores.
\begin{align}
  \text{Excitation Ratio} &= \frac{\text{Max Longitudinal Excitation}}{\text{Max Transverse Excitation}}\\
                          &= \frac{\eta_{\text{exc}}(\Theta = 0)}{\eta_{\text{exc}}(\Theta = \pi/2, \Phi = \phi_{\text{pol}})}\\
                          &= \frac{x^2\cos^2\phi_{\text{pol}} + \frac{1}{4}w^2(z)\sin^2\phi_{\text{pol}}}{z_0^2}\label{eq:ratio}
\end{align}

At first glance equation \ref{eq:ratio} looks bleak---the excitation ratio grows
without bound in the $x$ direction which means that the longitudinal excitation
becomes a larger fraction of the total excitation as we move away from the plane
of the light sheet. Fortunately, we only care about regions of the beam with a
high intensity, so we look at the intensity weighted excitation ratio instead.
\begin{align*}
    \text{Intensity Weighted Excitation Ratio} &= \frac{w_0}{w(z)}e^{-\frac{2x^2}{w^2(z)}}\frac{x^2\cos^2\phi_{\text{pol}} + \frac{1}{4}w^2(z)\sin^2\phi_{\text{pol}}}{z_0^2}\label{eq:intratio} \\
\end{align*}
We can interpret the intensity weighted excitation ratio as the fraction of the
maximum signal (created by a transverse fluorophore at the origin) that we
ignore by ignoring longitudinal excitation. Figure \ref{fig:error} shows that
the intensity weighted excitation ratio is <2\% for the imaging parameters used
in Wu et. al. \cite{wu2013}.

\fig{figures/error.pdf}{1.0}{Intensity weighted excitation ratio as a
  function of position using the parameters in Wu et. al. \cite{wu2013}:
  $w_0 = 1.2\ \mu$m, $z_0 = 9\ \mu$m, $\lambda = 488$ nm, FOV$\approx$ 80$\times$80
  $\mu$m${}^2$. The maximum values are at the edge of the FOV and are <2\%.}{error}

As a rough heuristic, if the fraction of the signal from noise and background is
greater than the intensity weighted excitation ratio we can justifiably ignore
the longitudinal component.

Figure 2 shows the maximum intensity weighted excitation ratio for an
80$\times$80 $\mu$m${}^2$ FOV as a function of $w_0$. The maximum intensity
weighted excitation ratio decreases as $w_0$ increases because wider beams
have smaller longitudinal components.

\fig{figures/max-error.pdf}{0.5}{Maximum intensity weighted excitation ration (see equation \ref{eq:intratio}) for an 80$\times$80 $\mu$m${}^2$ FOV with $\lambda = 488$ nm.}{max-error}

@@

To ignore the longitudinal component, we require that the fraction of the signal
from noise and background is greater than the intensity weighted excitation
ratio (<2\% with current imaging parameters). We'll need to be careful about
longitudinal excitation if we want to use beams that are more strongly focused.
